---
title: "title"
author: "name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_download: true
    code_folding: hide
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r, include=FALSE}

# settings, placed in a chunk that will not show in the .html file (because include=FALSE) 

# do not print messages and warnings to the html file
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

# disables scientific notation so that small numbers appear as eg "0.00001" rather than "1e-05"
options(scipen = 999) 

```

# Level of heading 1

## Level of heading 2

### Level of heading 3

Text

```{r}

# code chunk

```

# License and citation

(c) [name] ([date])

All text is [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/deed.en): you can use or modify it as you please as long as you cite its source.

All code is [MIT](https://opensource.org/license/mit) licensed: you can use or modify it as you please.

Suggested citation:

[add] 

# Session info

You can help make your code and results more reproducible by including a `sessionInfo()` call at the end of your scripts. This prints details of your operating system, R version, and packages that were used in your script. When you click knit to create a

```{r}

sessionInfo()

```
