---
title: "Data types - in-class exercises"
author: "Ian Hussey"
date: today
editor: source
format:
  html:
    theme:
      light: flatly
      dark: darkly
    toc: true
    toc-location: right
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-tools: true
    code-copy: true
    code-link: true
    code-overflow: wrap
    df-print: paged
    embed-resources: true
    fig-width: 7
    fig-height: 5
    include-in-header:
      text: |
        <style>
        /* make quarto's light/dark toggle visible in standalone documents */
        .quarto-color-scheme-toggle.top-right {
          position: fixed; top: 1rem; left: 1rem; right: auto; z-index: 1050;
          display: flex; align-items: center; gap: .4rem;
          padding: .3rem .7rem; border: 1px solid #adb5bd; border-radius: .5rem;
          background-color: var(--bs-body-bg); color: var(--bs-body-color);
          font-size: .9rem; text-decoration: none;
        }
        .quarto-color-scheme-toggle.top-right .bi::before { width: 1.4rem; height: 1.4rem; background-size: 1.4rem 1.4rem; }
        body.quarto-light .quarto-color-scheme-toggle.top-right .bi::before { filter: brightness(.4); }
        .quarto-color-scheme-toggle.top-right::after { content: "Dark mode"; }
        body.quarto-dark .quarto-color-scheme-toggle.top-right::after { content: "Light mode"; }
        body { padding-top: 3rem; } /* room for the toggle above the title */
        </style>
execute:
  message: false
  warning: false
---

Put this file in `exercises/`!

# Dependencies

```{r}

library(dplyr) # for %>%, mutate, filter, summarize, group_by, if_else, case_when, near, between
library(readr) # for read_csv
library(tibble) # for tribble
library(roundwork) # for round_up
library(truffle) # for round_p_value

```

# Today's data

Run this chunk, then look at the data frames.

`dat_survey` is read from a small .csv-style text, so that `read_csv()` has to guess each column's type, just as it does with real data files. Missing ages were coded as `999`.

`dat_rt` contains trial-level data from a reaction time task.

`dat_results` contains the results of some hypothesis tests.

```{r}

dat_survey <- read_csv(I("id,age,consent,mean_rt,comments
1,23,yes,512.5,
2,twenty,yes,498.25,
3,999,no,,
4,,yes,601,
5,35,yes,550.75,"),
show_col_types = FALSE)

dat_rt <- tribble(
  ~id, ~trial, ~rt, ~correct,
  1,   1,      512, 1,
  1,   2,      604, 1,
  1,   3,      488, 0,
  2,   1,      701, 0,
  2,   2,      655, 0,
  2,   3,      690, 0,
  3,   1,      530, 1,
  3,   2,      575, 1,
  3,   3,      549, 1
)

dat_results <- tribble(
  ~hypothesis, ~estimate, ~p,
  "H1",        0.2345,    0.2346,
  "H2",        0.1250,    0.0340,
  "H3",        0.5125,    0.0000004,
  "H4",        0.0875,    0.04999
)

```

# Types

## What types did `read_csv()` guess?

Use `glimpse()` to see the type of each column in `dat_survey`. For each column, explain why `read_csv()` chose that type. Which type would you expect for 'comments'? Why?

```{r}



```

## Predict before you run

Before running each line: what will the result be, and what type?

```{r}

c(1, 2, "3")
c(TRUE, FALSE, 2)
c(1, NA, 3)
as.numeric(c("23", "twenty", "999"))

```

## Fix the age column

Clean the 'age' column of `dat_survey` so that it is numeric, the missing data code is `NA`, and "twenty" is converted to 20 rather than lost. Check how many ages are `NA` before and after your cleaning.

```{r}



```

# `NA`

## Predict before you run

Before running each line: what will the result be?

```{r}

NA > 5
NA == NA
sum(c(1, 2, NA))
mean(c(NA, NA), na.rm = TRUE)
NA & FALSE

```

## A filter that drops too much

The code below is meant to remove participants with the missing data code `999`. How many rows do you expect it to return? How many does it return? Why? Fix it.

```{r}

dat_survey %>%
  mutate(age = as.numeric(age)) %>%
  filter(age != 999)

```

```{r}



```

## Typed `NA`s

Which of these chunks will throw an error? Why? Fix the one(s) that do.

```{r}
#| error: true

dat_survey %>%
  mutate(consent = case_when(consent == "yes" ~ TRUE,
                             consent == "no" ~ FALSE,
                             TRUE ~ NA))

```

```{r}
#| error: true

dat_survey %>%
  mutate(comments = if_else(id == 3, "withdrew consent", NA_real_))

```

```{r}
#| error: true

dat_survey %>%
  mutate(comments = if_else(id == 3, "withdrew consent", comments))

```

## Summaries when everything is missing

Calculate each participant's mean RT on correct trials only in `dat_rt`, along with the number of correct trials. Hint: first create a column containing the RT on correct trials and `NA` on error trials.

What do you get for participant 2? Why? Modify your code so that participant 2 gets `NA` instead.

```{r}



```

# Floating point

## Predict before you run

Before running each line: TRUE or FALSE?

```{r}

0.1 * 3 == 0.3
1 - 0.4 == 0.6
near(0.1 * 3, 0.3)
(1 - 0.95) < 0.05

```

Why do some of these give the "wrong" answer? How would you check what number is really stored?

## Proportions

Calculate each participant's proportion of correct trials in `dat_rt`. Then filter for participants who made exactly one error, i.e., whose proportion correct is `1 - 1/3`. Do it with `==` and with `near()`. Do they give the same result? Why?

```{r}



```

## Checking ranges with `between()`

Use `between()` to create a column in `dat_rt` that is TRUE if the RT is between 200 and 600 ms (inclusive).

```{r}



```

# Rounding and reporting

## Predict before you run

Before running: what will each of these return?

```{r}

round(2.5)
round(0.125, digits = 2)
round_up(2.5)
round_up(0.125, digits = 2)

```

## Spot the bug

This code is meant to flag significant results and report APA-style *p*-values. Run it. Which hypotheses are marked as significant? Is that correct? What went wrong?

```{r}

dat_results %>%
  mutate(p = round_p_value(p),
         significant = p < .05)

```

Fix it.

```{r}



```

## A results table

Using `dat_results`, create a table suitable for a manuscript:

1. A column `significant` that is TRUE if *p* < .05.
2. The estimate rounded to 2 decimal places using round-half-up.
3. The *p*-value in APA format.

Which of these steps must come first? Why?

```{r}



```

# Fix the broken pipe

This code is meant to clean `dat_survey`, then calculate the mean age and mean RT, rounded to 2 decimal places. It has (at least) four bugs. Find and fix them one at a time.

```{r}
#| eval: false

dat_survey %>%
  mutate(age = if_else(age == "twenty", "20", age),
         age = as.numeric(age),
         age = if_else(age == 999, NA_character_, age)) %>%
  summarize(mean_age = mean(age),
            mean_rt = mean(mean_rt, na.rm = TRUE)) %>%
  mutate(across(everything(), round))

```

# Looking ahead

In this session, we converted "twenty" to 20 by hand. Imagine a data set with hundreds of ages written as words, e.g., "thirty five", "forty-two", "Twenty One". What tools would you need to convert them all? What problems might you run into?
