---
title: "Reproducible reports"
format:
  html:
    toc: true
    toc_float: true
    code-fold: show
    code-tools: true
    self-contained: true
---

```{r}
#| include: false

# settings, placed in a chunk that will not show in the .html file (because include=FALSE) 

# disables scientific notation so that small numbers appear as eg "0.00001" rather than "1e-05"
options(scipen = 999)  

```

## Literature programming and code chunks

Literate programming is the idea that code and text should be written in the same document to produce a narrative with reproducible results. It is therefore very suited to writing scientific reports and manuscripts.

Code can be written in 'in line' in the text as follows: `r 1+1`. In the RMarkdown document, you have hover over the in line code and press enter or return to run the code.

You can also write it in chunks:

```{r}

2+2

```

Output appears below chunks. You can run all code in a chunk by clicking the right-arrow button to the right of the chunk. You can also run all previous chunks in a document not including the current chunk by clicking the downward arrow button to the right of the chunk.

## Math via LaTeX

You can include math in line with LaTeX code placed between dollar signs: e.g., "$\eta_{p}^{2}$ = 0.03".

You can also write longer chunks of LaTeX, for example to specify that the mean ($\bar{x}$) is the sum of all elements of the vector $x$ divided by number of elements in the vector ($n$).

This code:

```         
$$
\bar{x} = \frac{1}{n} \sum_{i=1}^n x_i.
$$
```

Produces this math:

$$
\bar{x} = \frac{1}{n} \sum_{i=1}^n x_i.
$$

## Source versus Visual editor

## kniting/rendering and reproducibilty

\TODO

## Levels of heading

\TODO
